manuscript_outline

### 2025-06-16 docs/manuscript/paper.Rmd
copy edit

| Class | Order | Genera | Species | Study |
|:---|:---|:---|:---|:---|
| Amphibia | Anura | *Lithobates* | *L. sylvaticus* | @moore1939 |
|  |  |  | *L. pipiens* |  |
|  |  |  | *L. clamitans* |  |
|  |  |  | *L. palustris* |  |
|  |  | *Ascaphus* | *A. truei* | @brown1975 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Reptilia | Squamata | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  |  | *Podarcis* | *P. mural*is | @vandamme1992 |
|  | Testudines | *Mauremys* | M. reevesii | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |
| Insecta | Plecoptera | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
|  | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  |  |  | 18 species | Developmental equations in @pritchard1987 |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |

@@

We evaluated eight candidate linear models that varied in the inclusion and interaction of year, COMID, and mountain range effects. Based on adjusted R², AICc, and RMSE, the best-supported model included year and COMID as additive fixed effects: days_to_hatch ~ year + comid.

@@

This model explained ~`r round(perf[1,4], 2)*100`% of the variation in developmental timing (adjusted R² = `r round(perf[1,4], 3)`), with year having a significant weak and negative effect (`r round(params[2,2], 2)` days/year, p = 0.001), indicating an overall trend toward faster development consistent with regional warming (Figure \@ref(fig:frog-preds)A). COMID effects captured the persistent differences in thermal regimes among streams, with estimated marginal means ranging from ~15 to ~44 days to hatch across sites (Figure \@ref(fig:frog-preds)B).

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD ~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD < 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast potential shifts under shifting climate patterns.

@@

Here, we used data from a common garden experiment on cabbage beetles (Colaphellus bowringi) reared at six constant temperatures (16–28 °C), spanning populations collected along a ~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, developmental rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes—where temperatures are generally warmer—developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].

>>>>

| Class | Order | Genera | Species | Study |
|:--------------|:--------------|:--------------|:--------------|:--------------|
| Amphibia | Anura | *Lithobates* | *L. sylvaticus* | @moore1939 |
|  |  |  | *L. pipiens* |  |
|  |  |  | *L. clamitans* |  |
|  |  |  | *L. palustris* |  |
|  |  | *Ascaphus* | *A. truei* | @brown1975 |
|  | Urodela | *Ambystoma* | *A. gracile* | @brown1976 |
| Reptilia | Squamata | *Sceloporus* | *S. undulatus* | @angilletta2000 |
|  |  | *Podarcis* | *P. mural*is | @vandamme1992 |
|  | Testudines | *Mauremys* | M. reevesii | @du2007 |
|  |  |  | 181 species | 141 studies in @while2018 |
| Insecta | Plecoptera | *Nemurella* | *N. pictetii* | @brittain1978, @elliott1984 |
|  |  | *Capnia* | *C. atra* | @brittain1984 |
|  |  | *Capnia* | *C. bifrons* | @elliott1986 |
|  |  | *Mesocapnia* | *M. oenone* | @brittain1984 |
|  |  | *Taeniopteryx* | *T. nebulosa* | @brittain1977 |
|  | Coleoptera | *Colaphellus* | *C. bowringi* | @tang2017 |
|  |  |  | 18 species | Developmental equations in @pritchard1987 |
| Malacostraca | Decapoda | *Pontastacus* | *P. leptodactylus* | @aydin2004 |

@@

We evaluated eight candidate linear models that varied in the inclusion and interaction of year, COMID, and mountain range effects. Based on adjusted R², AICc, and RMSE, the best-supported model included year and COMID as additive fixed effects: days_to_hatch \~ year + comid.

@@

This model explained \~`r round(perf[1,4], 2)*100`% of the variation in developmental timing (adjusted R² = `r round(perf[1,4], 3)`), with year having a significant weak and negative effect (`r round(params[2,2], 2)` days/year, p = 0.001), indicating an overall trend toward faster development consistent with regional warming (Figure \@ref(fig:frog-preds)A). COMID effects captured the persistent differences in thermal regimes among streams, with estimated marginal means ranging from \~15 to \~44 days to hatch across sites (Figure \@ref(fig:frog-preds)B).

@@

Our results indicate developmental rates were highly variable over different physical habitats (streams originating in lowland hills compared with those originating in mountain ranges with permanent snow fields), with sites in Olympic National Park consistently having the longest development times and the greatest interannual variability (SD \~ 11 days), while lowland streams in the Willapa Hills developed fastest with minimal variability (SD \< 3 days). These results demonstrate the capacity of **hatchR** to detect spatiotemporal phenological variation and forecast potential shifts under shifting climate patterns.

@@

Here, we used data from a common garden experiment on cabbage beetles (*Colaphellus bowringi*) reared at six constant temperatures (16–28 °C), spanning populations collected along a \~3,440 km latitudinal gradient in China [@tang2017]. Although all beetles were reared under identical conditions, developmental rates differed among populations, particularly at cooler temperatures. Populations from southern latitudes—where temperatures are generally warmer—developed more quickly than northern populations, a pattern consistent with cogradient variation, where genetic and environmental effects on a phenotype act in the same direction [@conover2009; @sparks2022].
